### 2018-07-13 submission/manuscript.tex
update manuscirpt order and results order (rough template

\textbf{\emph{Mock community relative abundance differs by GC content.}}
There was a significant difference in relative abundance between
high/low GC content based on either the V4 16S rRNA gene region or the
whole genome {[}Figure 1 \& Table S1{]}. However, using only the GC
content of the V4 region resulted in less differences than when using
the whole genome {[}Table S1{]}. The differences between high and low GC
content groups were significant as early as 20 cycles in all polymerases
except Q5 {[}Table S1{]}. The highest (\emph{Staphylococcus}) and lowest
(\emph{Pseudomonas}) GC content bacteria were the most divergent from
the expected relative abundance of 0.12 {[}Figure 1{]}. These
observations suggest that the high fidelity DNA polymerases tested do
not prevent GC-based amplification error.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used.}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 2A \& Table S2{]}.
Thus, the majority of the differences observed across the number of
cycles was between Kappa and other polymerases {[}Figure 2A \& Table
S3{]}. The total sequences that contained at least one error was also
polymerase dependent, with the majority of differences being between
Kappa or Accuprime and the other polymerases {[}Table S2 \& S3{]}. These
differences in error rates were not due to polymerase dependent
differences in base substitution rate {[}Figure S1{]}. Collectively, the
results suggest that sequence error is dependent on polymerase, persists
across the number of cycles used, and are not due to any bias towards a
specific base substitution.

\textbf{\emph{Prevalence of chimeric sequences are polymerase dependent
and correlate with the number of OTUs.}} Based on the previous results,
we examined whether chimeras also were dependent on polymerase and
whether this could affect the total number of OTUs. We observed
significant differences in the chimera prevalence based on polymerase
across the number of cycles used (P-value \textless{} 0.05) {[}Table
S2{]}. Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
plateau in the total percent of chimeras that were removed that was
similar for all polymerases {[}Figure 2D{]}. A positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 2E{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms that the
number of OTUs is dependent on the prevalence of these chimeric
sequences.

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used.}} In addition to the sequence error and chimera differences
between polymerases, we also investigated whether there were polymerase
or number of cycle dependent effects on the observed number of OTUs.
There was a polymerase dependent difference in the range of the number
of OTUs in the mock community samples {[}Figure 3A{]}. The closest any
of the polymerases came to a total of 8 OTUs, created by running the
mock reference 16S sequences through mothur processing, was at 25 and 30
cycles {[}Figure 3A{]}. Differences between polymerases for the number
of OTUs created were observed as early as 20 cycles in the mock
community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
Tukey post-hoc test, the majority of differences for the number of OTUs
detected in the mock community was largely due to Kappa and Platinum
versus the other polymerases across the number of cycles used {[}Table
S5{]}. Similarly, differences in the range of the number of OTUs
detected for fecal samples was dependent on the polymerase used
(e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion at 35 cycles = 84
- 136) {[}Figure 3B{]}. There was also a trend for lower number of
cycles (15-20) to result in a reduced range in the number of OTUs versus
higher number of cycles (25, 30, and 35) for all polymerases
(e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35 cycles = 84 -
136) {[}Figure 3B{]}. Unlike mock community samples, an overall
difference in the number of OTUs was only detected between polymerases
at 35 cycles in fecal samples (F-stat \textgreater{} 16.35, P-value =
9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to identify
any differences between specific polymerase groups (P-value
\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
noticeably larger difference between polymerases in the number of OTUs
than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
consistent across the number of amplification cycles tested {[}Figure
3{]}. Based on these observations, it is clear that larger differences
in total number of OTUs based on polymerases are found in mock versus
fecal communities.

\textbf{\emph{The bacterial communities generated by the polymerases
were similar but varied by number of cycles.}} Next, we investigated the
effect polymerases and number of cycles had on the bacterial community
composition in more detail by using the Bray-Curtis index. We observed
no difference in the bacterial community composition between any measure
tested when using the mock community (P-value \textgreater{} 0.05)
{[}Figure 4A{]}. With respect to fecal samples, within each respective
5-cycle increment comparison there was no difference in Bray-Curtis
index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
4B{]}. Using PERMANOVA to test for community differences based on any of
the number of cycles within polymerases, only Phusion had cycle
dependent differences (P-value = 0.03). Additionally, Phusion was one of
two polymerases that had enough sequences to be rarefied to 1000
sequences at 15 cycles. There was an overall decrease in Bray-Curtis
index by cycle comparison group (e.g.~the 15 cycle versus 20 cycle group
compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus
20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
(P-value \textless{} 0.05). Overall, these data suggest that the number
of cycles can change the bacterial community independent of the
polymerase used to generate the sequences.

\textbf{\emph{Polymerase dependent differences in taxa are more
pronounced in mock versus fecal samples.}} Finding minimal differences
in overall bacterial community composition, we next examined if specific
taxa vary depending on polymerase and number of cycles. Further
investigation of mock community samples found that OTUs taxonomically
classified to \emph{Salmonella}, \emph{Escherichia},
\emph{Enterococcus}, and \emph{Staphylococcus} differed based on
polymerase used {[}Table S6 \& S7{]}. These specific OTUs were all
members of the mock community and the differences were most pronounced
at 25 and 30 cycles {[}Table S7{]}. The majority of the differences
between taxa were due to differences between Kappa and the other
polymerases {[}Table S7{]}. When differences in OTUs based on polymerase
were investigated for fecal samples, no differences could be found at
any number of cycles (P-value \(\geqslant\) 0.01, corrected P-value =
1). These results suggest that taxa specific differences in mock
community samples do not carry over to actual fecal community samples.

\textbf{\emph{Variation due to polymerase and number of cycles is
dwarfed by the biological variation of an individual.}} Using 16S rRNA
gene sequencing data there was a large disparity in a model's ability to
correctly classify samples based on polymerase used, number of cycles,
and individual {[}Figure 5A{]}. The model classification success for an
individual was significantly better than all other models tested
{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most important
variables to the polymerase and number of cycle models were not the most
important variables in the model classifying individuals {[}Figure
5B{]}. This observation shows that OTUs that change due to the number of
cycles and polymerase used are not the same as OTUs that differentiate
between individuals {[}Figure 5B{]}. Although there are a variety of
errors associated with the amplification process, they are smaller than
the bacterial community variation between individuals.

>>>>

\textbf{\emph{Error differences due to number of cycles or polymerase
used are detectable in mock communities.}} There was a significant
difference in relative abundance between high/low GC content based on
either the V4 16S rRNA gene region or the whole genome {[}Figure 1 \&
Table S1{]}. However, using only the GC content of the V4 region
resulted in less differences than when using the whole genome {[}Table
S1{]}. The differences between high and low GC content groups were
significant as early as 20 cycles in all polymerases except Q5 {[}Table
S1{]}. The highest (\emph{Staphylococcus}) and lowest
(\emph{Pseudomonas}) GC content bacteria were the most divergent from
the expected relative abundance of 0.12 {[}Figure 1{]}. These
observations suggest that the high fidelity DNA polymerases tested do
not prevent GC-based amplification error.

The median error rate varied by polymerase, with Kappa having the
highest error rate of all the polymerases across the number of cycles
{[}Figure 2A \& Table S2{]}. Thus, the majority of the differences
observed across the number of cycles was between Kappa and other
polymerases {[}Figure 2A \& Table S3{]}. The total sequences that
contained at least one error was also polymerase dependent, with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S2 \& S3{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific base substitution.

Based on the previous results, we examined whether chimeras also were
dependent on polymerase and whether this could affect the total number
of OTUs. We observed significant differences in the chimera prevalence
based on polymerase across the number of cycles used (P-value
\textless{} 0.05) {[}Table S2{]}. Differences in chimera prevalence
between Platinum and all other polymerases accounted for the majority of
these differences {[}Table S3{]}. Accuprime\textsuperscript{TM} had the
lowest chimera prevalence of all polymerases regardless of whether
chimera removal with VSEARCH was used {[}Figure 2B \& 2C{]}. The number
of cycles used clearly increased chimera prevalence for all polymerases
but the rate of increase differed {[}Figure 2B \& 2C{]}. Additionally,
there was a plateau in the total percent of chimeras that were removed
that was similar for all polymerases {[}Figure 2D{]}. A positive
correlation was observed between chimeric sequences and the number of
OTUs for all polymerases {[}Figure 2E{]}. This positive correlation was
strongest for Accuprime\textsuperscript{TM}, Platinum, and Phusion
{[}Figure 2E{]}. This data suggests that chimera prevalence depends on
polymerase, is made worse by increasing the number of cycles, and
confirms that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

\textbf{\emph{Bacterial community composition differences based on
number of cycles or polymerase used are less obvious in mock community
samples.}} Due to the close relationship between chimera prevalence and
the number of OTUs, we also observed a polymerase dependent difference
in the range of the number of OTUs in the mock community samples based
on number of cycles and polymerase {[}Figure 3A{]}. The closest any of
the polymerases came to a total of 8 OTUs, created by running the mock
reference 16S sequences through mothur processing, was at 25 and 30
cycles {[}Figure 3A{]}. Differences between polymerases for the number
of OTUs created were observed as early as 20 cycles in the mock
community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
Tukey post-hoc test, the majority of differences for the number of OTUs
detected in the mock community was largely due to Kappa and Platinum
versus the other polymerases across the number of cycles used {[}Table
S5{]}.

Next, we investigated the effect polymerases and number of cycles had on
the bacterial community composition in more detail by using the
Bray-Curtis index. We observed no difference in the bacterial community
composition between any measure tested when using the mock community
(P-value \textgreater{} 0.05) {[}Figure 4A{]}. Finding minimal
differences in overall bacterial community composition, we next examined
if specific taxa vary depending on polymerase and number of cycles. We
found that OTUs that taxonomically classified to \emph{Salmonella},
\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
differed based on polymerase used {[}Table S6 \& S7{]}. These specific
OTUs were all members of the mock community and the differences were
most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
differences between taxa were due to differences between a single
polymerase (Kappa) and the other polymerases {[}Table S7{]}.

When assessing the Random Forest models there was a difference between
models built to classify number of cycles versus those built to classify
polymerase used {[}P-value = 5.06e-07{]}. Although there was a
difference, neither model was successful at correctly classifying
samples based on 16S rRNA gene sequencing data (cycles model,
probability of correct classification = 0.42 (0.4 - 0.43)(min - max)
versus polymerase model, probability of correct classification = 0.39
(0.37 - 0.4)). Our observations in mock communities suggest that
although differences are detected based on number of cycles or
polymerase used, the overall community-wide differences are subtle.

\textbf{\emph{Few differences based on number of cycles and polymerased
used were identified in fecal samples.}} Similarly, differences in the
range of the number of OTUs detected for fecal samples was dependent on
the polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus
Phusion at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend
for lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 3B{]}. Unlike mock community samples, an
overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerase groups (P-value
\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
noticeably larger difference between polymerases in the number of OTUs
than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
consistent across the number of amplification cycles tested {[}Figure
3{]}. Based on these observations, it is clear that larger differences
in total number of OTUs based on polymerases are found in mock versus
fecal communities.

With respect to fecal samples, within each respective 5-cycle increment
comparison there was no difference in Bray-Curtis index between the
polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}. Using
PERMANOVA to test for community differences based on any of the number
of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03). Additionally, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 sequences
at 15 cycles. There was an overall decrease in Bray-Curtis index by
cycle comparison group (e.g.~the 15 cycle versus 20 cycle group compared
to the 30 cycle versus 35 cycle group) (P-value \textless{} 0.01)
{[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus 20
cycle and 20 cycle versus 25 cycle comparison groups had a higher
Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
(P-value \textless{} 0.05). Overall, these data suggest that the number
of cycles can change the bacterial community independent of the
polymerase used to generate the sequences.

When differences in OTUs based on polymerase were investigated for fecal
samples, no differences could be found at any number of cycles (P-value
\(\geqslant\) 0.01, corrected P-value = 1). These results suggest that
taxa specific differences in mock community samples do not carry over to
actual fecal community samples.

Using 16S rRNA gene sequencing data there was a large disparity in a
model's ability to correctly classify samples based on polymerase used,
number of cycles, and individual {[}Figure 5A{]}. The model
classification success for an individual was significantly better than
all other models tested {[}Figure 5A \& Table S8{]}. Additionally, the
top 10 most important variables to the polymerase and number of cycle
models were not the most important variables in the model classifying
individuals {[}Figure 5B{]}. This observation shows that OTUs that
change due to the number of cycles and polymerase used are not the same
as OTUs that differentiate between individuals {[}Figure 5B{]}. Although
there are a variety of errors associated with the amplification process,
they are smaller than the bacterial community variation between
individuals.
